---
title: "Cash Flow Matching with Linear Programming"
output:  
      html_document:  
        keep_md: true 
---

Dedication or cash-flow matching is a technique used to fund known liabilities in
the future. The intent is to form a portfolio of assets whose cash inflows will exactly
match cash outflows of the liabilities. The liabilities will be paid off, as they become
due, without the need to sell or buy assets in the future. Dedicated portfolios usually only
consist of risk-free non-callable bonds since future cash inflows need to be known
when the portfolio is constructed. In this project, we will be writing and using an 
R program to construct a dedicated portfolio.

Given the liability schedule and bonds shown below, we will now formulate the problem as a linear program and solve it using lpSolveAPI.

#####Formulate the dedicated portfolio construction problem as a linear program  
**Decision variables:** (x1), ... , (x10) amount of each bond  
                    (z1), ... , (z7) excess cash at the end of each year (excluding the excess cash at the                                         last period)  

**Minimize:** 102(x1) + 99(x2) + 101(x3) + 98(x4) + 98(x5) + 104(x6) + 100(x7) + 101(x8) + 102(x9) + 94(x10)
 cost of each bond

**Subject To:**  
 year 1: 12000 = 100(x1) + 5(x1) + 3.5(x2) + 5(x3) + 3.5(x4) + 4(x5) + 9(x6) + 6(x7) + 8(x8) + 9(x9) + 7(x10) - (z1)  
 year 2: ...  

**Variables** = bond terms (10), excess cash (7)  
**Constraints** = years (8)  

######Set parameters for example portfolio
```{r}
library(lpSolveAPI)

p = c(102,99,101,98,98,104,100,101,102,94) #price
c = c(5,3.5,5,3.5,4,9,6,8,9,7) #coupon
m = c(1,2,2,3,4,5,5,6,7,8) #maturity
l = c(12000,18000,20000,20000,16000,15000,12000,10000) #liabilities (RHS)
```

######Code for solving optimization
```{r}
num.constraints = length(l) #set num of constraints to length of liabilities
num.bonds = length(c) #set num of bonds to length of coupons

my.lp = make.lp(num.constraints, num.bonds+num.constraints-1) #constraints, variables
check = matrix(0, num.constraints, num.bonds+num.constraints-1) #matrix to check that values are correctly inputted

z.mat = diag(-1, num.constraints, num.constraints-1) #initialize z mat and assign -1 to current year excess cash
z.mat[col(z.mat) == row(z.mat)-1] = 1 #assign 1 to previous year excess cash
z = 1
for(i in (num.bonds+1):(num.bonds+num.constraints-1)) { #loop to add z's to columns
  set.column(my.lp, i, z.mat[,z])
  check[,i] = z.mat[,z]
  z = z + 1
}

for(bond in (1:num.bonds)){ #loop to add columns for x variables
  maturity = m[bond] #set maturity to year the bond matures
  col = rep(0, num.constraints) #assign all variables 0's
  col[1:maturity] = c[bond] #assign coupons for all eligible years
  col[maturity] = (100 + c[bond]) #assign 100 + coupon value at maturity
  set.column(my.lp, bond, col) #set column
  check[, bond] = col
}

set.objfn(my.lp, c(p, rep(0, num.constraints-1)))
set.constr.type(my.lp, rep("=", num.constraints))
set.rhs(my.lp, l)
```

######Solve
```{r}
solve(my.lp)
```
  
######Get optimal variables
```{r}
get.variables(my.lp)
```

    
Using the optimization code above, we can write a function that can construct a portfolio for any set of liabilities and bonds.

######Function that takes in bond portfolio parameters and returns a .lp object
```{r}
dedicate_g1 = function (p,c,m,l) {
  num.constraints = length(l) #set num of constraints to length of liabilities
  num.bonds = length(c) #set num of bonds to length of coupons
  
  my.lp = make.lp(num.constraints, num.bonds+num.constraints-1) #constraints, variables
  check = matrix(0, num.constraints, num.bonds+num.constraints-1) #matrix to check that values are correctly inputted
  
  z.mat = diag(-1, num.constraints, num.constraints-1) #initialize z mat and assign -1 to current year excess cash
  z.mat[col(z.mat) == row(z.mat)-1] = 1 #assign 1 to previous year excess cash
  z = 1
  for(i in (num.bonds+1):(num.bonds+num.constraints-1)) { #loop to add z's to columns
    set.column(my.lp, i, z.mat[,z])
    check[,i] = z.mat[,z]
    z = z + 1
  }
  
  for(bond in (1:num.bonds)){ #loop to add columns for x variables
    maturity = m[bond] #set maturity to year the bond matures
    col = rep(0, num.constraints) #assign all variables 0's
    col[1:maturity] = c[bond] #assign coupons for all eligible years
    col[maturity] = (100 + c[bond]) #assign 100 + coupon value at maturity
    set.column(my.lp, bond, col) #set column
    check[, bond] = col
  }
  
  set.objfn(my.lp, c(p, rep(0, num.constraints-1)))
  set.constr.type(my.lp, rep("=", num.constraints))
  set.rhs(my.lp, l)
  return(my.lp)
}
```

######Test function on example portfolio
```{r}
old_port = dedicate_g1(p,c,m,l)
solve(old_port)
```

```{r}
get.variables(old_port)
#looks right
```

```{r}
get.objective(old_port)
#looks right
```

  
Next, we will construct a dedicated portfolio using the liability stream below and current bond information from [The Wall Street Journal (WSJ) Online U.S. Treasury Quotes](http://online.wsj.com/mdc/public/page/2_3020-treasury.html). We will have to account for the fact that coupons for these bonds are paid semi-annually.

#####Read in file with current bond information  
```{r}
bonds = read.csv('bonds.csv', header=TRUE) # csv containing maturity date, coupon, and ask price provided on the website
```

######Create a column of date objects
```{r}
bonds$Matur_Date = as.Date(as.character(bonds$Maturity),format="%m/%d/%Y")
```

######Remove all bonds maturing after last liability date
```{r}
bonds = bonds[bonds$Matur_Date <= as.Date('12/31/2021',format="%m/%d/%Y"),]
```

######Create a list of liability dates
```{r}
liability_dates = c('6/30/2016','12/31/2016','6/30/2017','12/31/2017','6/30/2018','12/31/2018','6/30/2019',
                    '12/31/2019','6/30/2020','12/31/2020','6/30/2021','12/31/2021')
liability_dates = as.Date(liability_dates,format="%m/%d/%Y")
```

######Assign maturity periods using the date lists
Goal is to have list of maturity due periods (numbering 1-12), similar to format given in example portfolio
```{r}
m = 0
for (i in 1:length(bonds$Matur_Date)) {
  for (j in 1:length(liability_dates)) {
    if (bonds$Matur_Date[i] <= liability_dates[j]) {
      m[i] = j
      break
    }
  }
}
```

######Use function to optimize new portfolio
```{r}
p = bonds$Asked #price
c = bonds$Coupon #coupon
l = c(9000000,9000000,10000000,10000000,6000000,6000000,9000000,9000000,10000000,10000000,5000000,3000000) #liabilities (RHS)

new_port = dedicate_g1(p,c,m,l)
solve(new_port)
```

The matrix below specifies how much and of each bond the portfolio should consist of. The last 11 variables are the z variables which represent excess cash carried over each year. They indicate that approximately $4.7M should be carried over from period 10 to 11.
```{r}
get.variables(new_port)
```

The cost of the portfolio is $74,042,228, which equates to $21,957,772 in savings (ignoring interest/time value of money).
```{r}
get.objective(new_port)
```


Finally, we will plot and interpret the sensitivity parameters of the dedicated portfolio.

######Plot bond and liability duals versus libability period (analagous to date)
```{r}

duals = get.dual.solution(new_port)
sensitivity = get.sensitivity.rhs(new_port)
dual_df = data.frame(sensitivity$duals,sensitivity$dualsfrom,sensitivity$dualstill)

liab_duals = dual_df[1:12,]
bond_duals = dual_df[13:236,]

```

```{r}
plot(m, bond_duals$sensitivity.duals, main="Bond Duals by Maturity Period")
```

The graph above shows the shadow prices for each of the 224 bonds by maturity period. The shadow price represents the increase in the objective value (cost of the portfolio) from a purchase of an additional bond. We see a positive relationship between maturity period and the duals, such that buying additional bonds with later maturity dates has a greater impact on the total portfolio cost than buying bonds with shorter term maturities.

This makes sense, because the further into the future a bond matures, the more coupons it will generate and the more expensive it is. For example, bond #1 has a maturity of 2/15/2016 and a shadow price of 8.05. This means buying an additional unit of bond 1 would increase the objective value by 8.05. Buying an additional bond that matures in a later period would increase the objective value by a greater amount.
```{r}
plot(c(1:12), liab_duals$sensitivity.duals, main="Liability Duals by Maturity Period")
```

Looking at liability duals by maturity shows a negative relationship. The first liability has a shadow price of 0.92, which means that an increase in the first liability by $1 requires a $0.92 increase in the objective value, or cost of the portfolio. Shadow prices fall as maturity dates rise because there is more time to accumulate return to cover the liability.