---
title: "SCO_Project1_Group1"
author: "Brian Lakey, Reshma Sekar, Julia Wu, Carlton Washburn"
date: "February 10, 2016"
---


##**PROBLEM 1**
####Formulate the dedicated portfolio construction problem as a linear program  

#### Setting up parameters for example portfolio  

**Decision variables:** (x1), ... , (x10) amount of each bond  
                    (z1), ... , (z7) excess cash at the end of each year (excluding the excess cash at the                                         last period)  

**Minimize:** 102(x1) + 99(x2) + 101(x3) + 98(x4) + 98(x5) + 104(x6) + 100(x7) + 101(x8) + 102(x9) + 94(x10)
 cost of each bond

**Subject To:**  
 year 1: 12000 = 100(x1) + 5(x1) + 3.5(x2) + 5(x3) + 3.5(x4) + 4(x5) + 9(x6) + 6(x7) + 8(x8) + 9(x9) + 7(x10) - (z1)  
 year 2: ...  

constraints = years (8)  
variables = bond terms (10), excess cash (7)  

####Setting up parameters for example portfolio
```{r}
library(lpSolveAPI)

p = c(102,99,101,98,98,104,100,101,102,94) #price
c = c(5,3.5,5,3.5,4,9,6,8,9,7) #coupon
m = c(1,2,2,3,4,5,5,6,7,8) #maturity
l = c(12000,18000,20000,20000,16000,15000,12000,10000) #liabilities (RHS)
```

####Setting up code for solving optimization
```{r}
num.constraints = length(l) #set num of constraints to length of liabilities
num.bonds = length(c) #set num of bonds to length of coupons

my.lp = make.lp(num.constraints, num.bonds+num.constraints-1) #constraints, variables
check = matrix(0, num.constraints, num.bonds+num.constraints-1) #matrix to check that values are correctly inputted

z.mat = diag(-1, num.constraints, num.constraints-1) #initialize z mat and assign -1 to current year excess cash
z.mat[col(z.mat) == row(z.mat)-1] = 1 #assign 1 to previous year excess cash
z = 1
for(i in (num.bonds+1):(num.bonds+num.constraints-1)) { #loop to add z's to columns
  set.column(my.lp, i, z.mat[,z])
  check[,i] = z.mat[,z]
  z = z + 1
}

for(bond in (1:num.bonds)){ #loop to add columns for x variables
  maturity = m[bond] #set maturity to year the bond matures
  col = rep(0, num.constraints) #assign all variables 0's
  col[1:maturity] = c[bond] #assign coupons for all eligible years
  col[maturity] = (100 + c[bond]) #assign 100 + coupon value at maturity
  set.column(my.lp, bond, col) #set column
  check[, bond] = col
}

set.objfn(my.lp, c(p, rep(0, num.constraints-1)))
set.constr.type(my.lp, rep("=", num.constraints))
set.rhs(my.lp, l)
```

  
##**PROBLEM 2**
####Solve the example portfolio case in R using lpSolveAPI() 
  
####Solving
```{r}
solve(my.lp)
```
  
####Getting optimal variables
```{r}
get.variables(my.lp)
```

    
##**PROBLEM 3**
#### Write a function in R that can construct a portfolio for any set of liabilities and bonds  

####Function takes in bond portfolio parameters and return a .lp object
```{r}
dedicate_g1 = function (p,c,m,l) {
  num.constraints = length(l) #set num of constraints to length of liabilities
  num.bonds = length(c) #set num of bonds to length of coupons
  
  my.lp = make.lp(num.constraints, num.bonds+num.constraints-1) #constraints, variables
  check = matrix(0, num.constraints, num.bonds+num.constraints-1) #matrix to check that values are correctly inputted
  
  z.mat = diag(-1, num.constraints, num.constraints-1) #initialize z mat and assign -1 to current year excess cash
  z.mat[col(z.mat) == row(z.mat)-1] = 1 #assign 1 to previous year excess cash
  z = 1
  for(i in (num.bonds+1):(num.bonds+num.constraints-1)) { #loop to add z's to columns
    set.column(my.lp, i, z.mat[,z])
    check[,i] = z.mat[,z]
    z = z + 1
  }
  
  for(bond in (1:num.bonds)){ #loop to add columns for x variables
    maturity = m[bond] #set maturity to year the bond matures
    col = rep(0, num.constraints) #assign all variables 0's
    col[1:maturity] = c[bond] #assign coupons for all eligible years
    col[maturity] = (100 + c[bond]) #assign 100 + coupon value at maturity
    set.column(my.lp, bond, col) #set column
    check[, bond] = col
  }
  
  set.objfn(my.lp, c(p, rep(0, num.constraints-1)))
  set.constr.type(my.lp, rep("=", num.constraints))
  set.rhs(my.lp, l)
  return(my.lp)
}
```

####Testing function on example portfolio
```{r}
old_port = dedicate_g1(p,c,m,l)
solve(old_port)
```

```{r}
get.variables(old_port)
#looks right
```

```{r}
get.objective(old_port)
#looks right
```

  
##**PROBLEM 4**
####Construct a dedicated portfolio using this liability stream and current bond information  

```{r}
bonds = read.csv('bonds.csv', header=TRUE) # we creted a csv document containing all of the important bond fields from the provided website
```

####Creating a column of date objects
```{r}
bonds$Matur_Date = as.Date(as.character(bonds$Maturity),format="%m/%d/%Y")
```

####Remove all bonds maturing after last liability date
```{r}
bonds = bonds[bonds$Matur_Date <= as.Date('12/31/2021',format="%m/%d/%Y"),]
```

####Creating a list of liability dates
```{r}
liability_dates = c('6/30/2016','12/31/2016','6/30/2017','12/31/2017','6/30/2018','12/31/2018','6/30/2019',
                    '12/31/2019','6/30/2020','12/31/2020','6/30/2021','12/31/2021')
liability_dates = as.Date(liability_dates,format="%m/%d/%Y")
```

####Assigning maturity periods using the date lists
Goal is to have list of maturity due periods (numbering 1-12), similar to format given in example portfolio
```{r}
m = 0
for (i in 1:length(bonds$Matur_Date)) {
  for (j in 1:length(liability_dates)) {
    if (bonds$Matur_Date[i] <= liability_dates[j]) {
      m[i] = j
      break
    }
  }
}
```

####Using function to optimize new portfolio
```{r}
p = bonds$Asked #price
c = bonds$Coupon #coupon
l = c(9000000,9000000,10000000,10000000,6000000,6000000,9000000,9000000,10000000,10000000,5000000,3000000) #liabilities (RHS)

new_port = dedicate_g1(p,c,m,l)
solve(new_port)
```

The chart below specifies how much and of which bonds the portfolio should invest in. The last 11 variables are excess cash carried over, indicating that approximately $4.7m should be carried over from period 10 to 11.
```{r}
get.variables(new_port)
```

Savings (below) of $21,957,772 (ignoring interest/time value of money)
```{r}
get.objective(new_port)
#sum(l) #pv of liabilities = $96,000,000
#optimal portfolio = $74,042,228
```

####Plotting bond and liability duals versus libability period (analagous to date)
```{r}

duals = get.dual.solution(new_port)
sensitivity = get.sensitivity.rhs(new_port)
dual_df = data.frame(sensitivity$duals,sensitivity$dualsfrom,sensitivity$dualstill)

liab_duals = dual_df[1:12,]
bond_duals = dual_df[13:235,]

```

The graph below shows the shadow prices for each of the 223 bonds by maturity period. The shadow price represents the increase in the objective value (cost of the portfolio) from a purchase of an additional bond. We see a positive relationship between maturity period and the duals, such that buying additional bonds with later maturity dates has a greater impact on the total portfolio cost than buying bonds with shorter term maturities.

This makes sense, because the further into the future a bond matures, the more coupons it will generate and the more expensive it is. For example, bond #1 has a maturity of 2/15/2016 and a shadow price of 8.05. This means buying an additional unit of bond 1 would increase the objective value by 8.05.  Buying an additional bond that matures in a later period would increase the objective value by a greater amount.
```{r}
plot(mm, bond_duals$sensitivity.duals, main="Bond Duals by Maturity Period")
```

Looking at liability duals by maturity shows a negative linear relationship. The first liability has a shadow price of 0.92, which means that an increase in the first liability by $1 requires a $0.92 increase in the objective value, or cost of the portfolio. Shadow prices fall as maturity dates rise because there is more time to accumulate return to cover the liability.
```{r}
plot(c(1:12), liab_duals$sensitivity.duals, main="Liability Duals by Maturity Period")
```